Characterizing Water Composition with an Autonomous Robotic Team Employing Comprehensive In Situ Sensing, Hyperspectral Imaging, Machine Learning, and Conformal Prediction

### 2024-03-10 paper/remotesensing-2870812-english.tex

The results of the model training procedure are presented in Table~\ref{tab:fit-results}. The performance of the model is identified by the $R^2$, RMSE, and MAE out-of-fold estimates (mean $\pm$ standard deviation) of the final hyperparameter-optimized model in the training set, with the target variables ranked in descending order by the $R^2$ value and separated by sensor type (physical variables, ions, biochemical variables, and chemical variables). The final hyperparameter values for each model are listed in Table~\ref{tab:hyperparameters} in Appendix~\ref{Model Hyperparameters}. The small variation in values across folds confirms that the reported performance is independent of how the training set was sampled. Furthermore, we report the interval width that yields a 90\% confidence interval on the holdout validation set determined by the conformal prediction procedure. We then evaluate how the estimated uncertainty generalizes by computing the \textit{\hl{empirical coverage}} on the holdout testing set: that is, we compute the percentage of predictions in the test set that \textit{\hl{actually}} fall within the estimated 90\% confidence interval.

@@

In \cite{robotTeam1}, we introduced this paradigm. In this study, we have built on this approach in three new ways. First, we have demonstrated the ability to effectively combine observations from disparate collections by augmenting the machine learning models with sufficient features describing the illumination and viewing geometries. As Figure~\ref{fig:downwelling-hist} indicates, we observed variation in the total downwelling intensity between the images collected on the same day \textit{\hl{and}} between each separate collection period. These within-collection variations are due to a combination of the stability of the UAV (on which the upward facing downwelling irradiance spectrometer is mounted) together with the occasional interference of clouds. Moreover, the assumption that the water's surface can be treated as Lambertian is clearly violated when the water is not perfectly still. Despite the potential impact of these limitations on the quality of the resulting reflectance data cubes, the smoothness of the maps generated by our models suggests that we have provided sufficient context by including the relevant solar illumination and viewing angles as additional features in the final dataset.  This fact is reinforced by the position of these variables as the most important features for \textit{each} of the estimated water quality variables. As long as we are primarily interested in these values and not the reflectances themselves, we are able to successfully account for these lighting effects when combining data from multiple collections.

@@

The third contribution of this work is the extension of our machine learning approach to enable uncertainty quantification through conformal prediction. For water quality risk assessment, the trustworthiness of model predictions is of equal or greater importance to the values themselves. However, robust uncertainty quantification has historically been challenging for many machine learning models, which behave like black boxes.
Conformal prediction is an attractive approach to enable model-agnostic uncertainty estimation and has recently seen adoption to remote sensing classification tasks such as land-type classification and object identification \cite{valle2023quantifying, zhu2024inductive}. In this setting, the goal is to produce predictive \textit{\hl{sets}} guaranteed to contain the correct class labels at a predetermined confidence level. Nevertheless, conformal prediction works equally well for regression tasks. By leveraging the large data volume collected by the robot team, we are able to simultaneously train predictive models and evaluate confidence intervals for their predictions. As the final column of Table~\ref{tab:fit-results} confirms, the empirical coverage on the holdout testing set provided by the inferred confidence intervals achieves the desired coverage to within 1\%. We chose to use a 90\% confidence interval for this study, but this can easily be adapted to suite the needs of a specific application if greater confidence is required.

Despite the wealth of information provided by the increased resolution of hyperspectral images, their considerable size impedes their \textit{\hl{complete}} utilization in real-time applications. Often, much of the available spectral information is discarded in favor of indices like the NDVI, which can be quickly computed as images are captured \cite{horstrand2019uav}. Utilizing machine learning allows us to take advantage of the full spectrum captured by each pixel while simultaneously reducing the size of the final data product to single-band ``images'' of selected water quality variables. We note that training a reduced-feature model without further hyperparameter optimization takes roughly one minute per target variable of interest using the processing computer included on the UAV. This means that, in principle, training data can be collected by the USV, imagery can be acquired and processed by the UAV, coincident records can be selected, and the resulting dataset can be used to train machine learning models all while investigators are still in the field. Analyzing the maps produced by applying each trained model enables areas of interest to be readily pinpointed, as demonstrated by the identification of slightly elevated levels of %Please ensure meaning has been retained. 
crude oil, optical brighteners, and CDOM in the eastern alcove of the pond on 23 November.

>>>>

The results of the model training procedure are presented in Table~\ref{tab:fit-results}. The performance of the model is identified by the $R^2$, RMSE, and MAE out-of-fold estimates (mean $\pm$ standard deviation) of the final hyperparameter-optimized model in the training set, with the target variables ranked in descending order by the $R^2$ value and separated by sensor type (physical variables, ions, biochemical variables, and chemical variables). The final hyperparameter values for each model are listed in Table~\ref{tab:hyperparameters} in Appendix~\ref{Model Hyperparameters}. The small variation in values across folds confirms that the reported performance is independent of how the training set was sampled. Furthermore, we report the interval width that yields a 90\% confidence interval on the holdout validation set determined by the conformal prediction procedure. We then evaluate how the estimated uncertainty generalizes by computing the empirical coverage on the holdout testing set: that is, we compute the percentage of predictions in the test set that actually fall within the estimated 90\% confidence interval.

@@

In \cite{robotTeam1}, we introduced this paradigm. In this study, we have built on this approach in three new ways. First, we have demonstrated the ability to effectively combine observations from disparate collections by augmenting the machine learning models with sufficient features describing the illumination and viewing geometries. As Figure~\ref{fig:downwelling-hist} indicates, we observed variation in the total downwelling intensity between the images collected on the same day and between each separate collection period. These within-collection variations are due to a combination of the stability of the UAV (on which the upward facing downwelling irradiance spectrometer is mounted) together with the occasional interference of clouds. Moreover, the assumption that the water's surface can be treated as Lambertian is clearly violated when the water is not perfectly still. Despite the potential impact of these limitations on the quality of the resulting reflectance data cubes, the smoothness of the maps generated by our models suggests that we have provided sufficient context by including the relevant solar illumination and viewing angles as additional features in the final dataset.  This fact is reinforced by the position of these variables as the most important features for each of the estimated water quality variables. As long as we are primarily interested in these values and not the reflectances themselves, we are able to successfully account for these lighting effects when combining data from multiple collections.

@@

The third contribution of this work is the extension of our machine learning approach to enable uncertainty quantification through conformal prediction. For water quality risk assessment, the trustworthiness of model predictions is of equal or greater importance to the values themselves. However, robust uncertainty quantification has historically been challenging for many machine learning models, which behave like black boxes.
Conformal prediction is an attractive approach to enable model-agnostic uncertainty estimation and has recently seen adoption to remote sensing classification tasks such as land-type classification and object identification \cite{valle2023quantifying, zhu2024inductive}. In this setting, the goal is to produce predictive sets guaranteed to contain the correct class labels at a predetermined confidence level. Nevertheless, conformal prediction works equally well for regression tasks. By leveraging the large data volume collected by the robot team, we are able to simultaneously train predictive models and evaluate confidence intervals for their predictions. As the final column of Table~\ref{tab:fit-results} confirms, the empirical coverage on the holdout testing set provided by the inferred confidence intervals achieves the desired coverage to within 1\%. We chose to use a 90\% confidence interval for this study, but this can easily be adapted to suite the needs of a specific application if greater confidence is required.

Despite the wealth of information provided by the increased resolution of hyperspectral images, their considerable size impedes their complete utilization in real-time applications. Often, much of the available spectral information is discarded in favor of indices like the NDVI, which can be quickly computed as images are captured \cite{horstrand2019uav}. Utilizing machine learning allows us to take advantage of the full spectrum captured by each pixel while simultaneously reducing the size of the final data product to single-band ``images'' of selected water quality variables. We note that training a reduced-feature model without further hyperparameter optimization takes roughly one minute per target variable of interest using the processing computer included on the UAV. This means that, in principle, training data can be collected by the USV, imagery can be acquired and processed by the UAV, coincident records can be selected, and the resulting dataset can be used to train machine learning models all while investigators are still in the field. Analyzing the maps produced by applying each trained model enables areas of interest to be readily pinpointed, as demonstrated by the identification of slightly elevated levels of %Please ensure meaning has been retained. 
crude oil, optical brighteners, and CDOM in the eastern alcove of the pond on 23 November.
